Reinforcement Learning Produces Dominant Strategies for the Iterated Prisoner's Dilemma

### 2017-10-03 plos/response_to_reviewers.tex

\section{Comments from second referee}

\begin{quote}
Honestly this work less dedicates to the field of evolutionary game from
scientific point of view. One point obviously pointing-out is that their
conclusion of “reinforcement learning produces dominant strategies to win the
Axelrod’s IPD tournament” is no surprising at all. It is because anyone can
agree that the more elaborate learning system, e.g. Structural Neural Network
System with brushing-up mechanism for weights like they did, can be
implemented, there might be emerging some smart and robust strategies gaining
high scores.  But, consequently I think the MS may be acceptable on Plos One
because it devotedly follows the historically important work by Axelrod, in
which quite few people still have been interested. However, the current MS
seems suffering from several crucial problems that must be definitely revised.
\end{quote}

\begin{quote}
1) The way of their description seems so verbose. But, simultaneously, some
descriptions, e.g. explaining models, seem little vague. More clear structure
and plain description should be expected. The MS structure should be rebuilt.
\end{quote}

We have trimmed various sections of the text and restructured the text to move
one of the sections to the Methods section. We feel the level of description of
each model is correct, furthermore it is accompanied by a diagram clearly
showing an example of each strategy. The results section is clearly written with
very little prose surrounding what is a standard scientific form for presenting
results. This is complemented by a verbose discussion section that is inline
with common scientific texts. Furthermore we note that the other reviewer did
not find a problem with our description. If the reviewer still feels that this
is not sufficient we would welcome a more detailed description of the issue that
persists.

### 2017-10-03 plos/main.tex
Trim some of the text.
This is in response to the second reviewer whose comment was rather
unhelpful. We should make a point about this to the editor.

The Prisoner's Dilemma (PD) is a two player game
used to model a variety of strategic interactions. Each player chooses
between cooperation (C) or defection (D). The payoffs of
the game are defined by the matrix $\begin{pmatrix} R & S \\ T & P
\end{pmatrix}$, where $T > R > P > S$ and $2R > T + S$. The PD is a one
round game, but is commonly studied in a manner where the prior outcomes
matter. This repeated form is called the Iterated Prisoner's
Dilemma (IPD).
The IPD is a common model in game theory,
frequently used  to understand the evolution of cooperative behaviour from complex
dynamics \cite{Axelrod1984}.

@@

Since Robert Axelrod's seminal tournament \cite{Axelrod1980a}, a number of
IPD\@ tournaments have been undertaken and are summarised in
Table~\ref{tbl:tournaments}. Further to the work described in
\cite{knight2016open} a regular set of standard, noisy~\cite{Bendor1991} and
probabilistic ending~\cite{Axelrod1980b} tournaments are carried out as more
strategies are added to the Axelrod library.
Details and results are available here:
\url{http://axelrod-tournament.readthedocs.io}. This work presents a detailed
analysis of a tournament with 176 strategies.

@@

Training of this strategy corresponds to finding maps from histories to
a probability of cooperation. The library includes a strategy trained
with $n_1 = m_1 = m_2 = 2$ that is \emph{mostly deterministic}, with 52 of the 64
probabilities being 0 or 1. At one time this strategy outperformed
EvolvedLookerUp2\_2\_2.

@@

Training of this strategy corresponds to finding parameters of the neural
network. An inner layer with just five nodes performs quite well in both deterministic and
noisy tournaments. The output of the ANN used in this work is deterministic;
a stochastic variant that outputs probabilities rather than exact moves could
be easily created.

@@

\begin{figure}[!hbtp]
    \centering
        \includegraphics[width=.5\textwidth]{FSM16.pdf}
        \caption{\bf Evolved\_FSM\_16: trained to maximise score in a standard
        tournament}
        \label{fig:fsm16}
\end{figure}

\begin{figure}
        \centering
        \includegraphics[width=.5\textwidth]{FSM16noise.pdf}
        \caption{Evolved\_FSM\_16\_Noise\_05: trained to maximise score in a
        noisy tournament}
        \label{fig:fsm16noise}
\end{figure}

@@

There are several strategies based on ensemble methods that
are common in machine learning called Meta strategies. These strategies are
composed of a team of other strategies. In each round, each member of the team
is polled for its desired next
move. The ensemble then selects the next move based on a rule, such as the
consensus vote in the case of MetaMajority or the best individual performance
in the case of MetaWinner. These strategies were among the best in the library
before the inclusion of those trained by reinforcement learning. The library
contains strategies containing teams of all the deterministic players, all the
memory-one players, and some others.

@@

The trained strategies (denoted by a \(^{*}\) in
Appendix~\ref{app:list_of_players}) were trained using reinforcement
learning algorithms. The ideas of reinforcement learning can be attributed to
the original work of \cite{turing1950computing} in which the notion that
computers would learn by taking random actions but according to a distribution
that picked actions with high rewards more often. The two particular algorithms
used here:

@@

Finally, Table~\ref{tbl:standard_ranks} and
Fig~\ref{fig:standard_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. Whilst there is some
stochasticity, the top three strategies almost always rank in the top three. For
example, the worst that the Evolved Lookerup 2 2 2 ranks in any tournament
is 8th.

@@

Noisy tournaments in which there is a 5\% chance that an action is
flipped are now described. As shown in Table~\ref{tbl:noisy_score} and
Fig~\ref{fig:noisy_score}, the best performing strategies in median payoff
are DBS, designed to account for noise, followed by two strategies trained in
the presence of noise and three trained strategies trained without noise. One of
the strategies trained with noise (PSO Gambler) actually performs less well than
some of the other high ranking strategies including
Spiteful TFT (TFT but defects indefinitely if the opponent defects twice
consecutively) and OmegaTFT (also designed to handle noise). While DBS is the clear
winner, it comes at a 6x increased run time over Evolved FSM 16 Noise 05.

@@

Fig~\ref{fig:comparison_cooperation_heatmaps_noisy_1}-~\ref{fig:comparison_cooperation_heatmaps_noisy_3} shows the rate of
cooperation in each round for the top three strategies (in the absense of noise)
and just as for the top performing strategies in the standard tournament
it is evident that
the strategies never defect first and learn to quickly punish poorer strategies.

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. An evolutionary algorithm produces strategies based on multiple
generic archetypes that are able to achieve a higher average
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies from the literature. Interestingly, a trained
memory one strategy produced by a particle swarm algorithm performs well, better
than human designed strategies such as Win Stay Lose Shift and zero determinant
strategies (which enforce a payoff difference rather than maximize total payoff).
The generic structure of the trained strategies did not appear to be
critical for the standard tournament -- strategies based on lookup tables,
finite state machines, neural networks, and stochastic variants all performed well.
Single layer neural networks performed well in both
noisy and standard tournaments
though these had some aspect of human involvement in the selection of features.
This is in line with the other strategies also where some human decisions are made
regarding the structure. For the LookerUp and Gambler archetypes
a decision has to be made
regarding the number of rounds of history and initial play that are to be used.
In contrast, the finite state machines and hidden Markov models
 required only a choice of the number
of states, and the training algorithm can eliminate unneeded states in the case
of finite state machines (evidenced by the unconnected nodes in the diagrams
for the included representations).

@@

In opposition to historical tournament results and community folklore,
our results show that complex strategies can be very effective for the
IPD\@. Designing complex strategies for the prisoner's dilemma appears to be
difficult for humans. Of all the human-designed
strategies in the library, only DBS consistently performs well, and it is
substantially more complex than traditional tournament winners like TFT, OmegaTFT,
and zero determinant strategies. Furthermore, dealing with noise is difficult
for most strategies. Two strategies designed specifically to account for noise,
DBS and OmegaTFT, perform well and only DBS performs better than the trained
strategies and \textbf{only} in some noisy contexts. Empirically we find that
DBS (with its default parameters) does not win tournaments at 1\% noise.
However DBS has a parameter that accounts for the expected amount of noise and a
followup study with various noise levels could make a more complete study of
the performance of DBS and strategies trained at various noise levels.

>>>>

The Prisoner's Dilemma (PD) is a two player game
used to model a variety of strategic interactions. Each player chooses
between cooperation (C) or defection (D). The payoffs of
the game are defined by the matrix $\begin{pmatrix} R & S \\ T & P
\end{pmatrix}$, where $T > R > P > S$ and $2R > T + S$. The PD is a one
round game, but is commonly studied in a manner where the prior outcomes
matter. This repeated form is called the Iterated Prisoner's
Dilemma (IPD).
The IPD is 
frequently used  to understand the evolution of cooperative behaviour from complex
dynamics~\cite{Axelrod1984}.

@@

Since Robert Axelrod's seminal tournament \cite{Axelrod1980a}, a number of
IPD\@ tournaments have been undertaken and are summarised in
Table~\ref{tbl:tournaments}. Further to the work described in
\cite{knight2016open} a regular set of standard, noisy~\cite{Bendor1991} and
probabilistic ending~\cite{Axelrod1980b} tournaments are carried out as more
strategies are added to the Axelrod library.
Details and results are available here:
\url{http://axelrod-tournament.readthedocs.io}. This work presents a detailed
analysis of tournaments with 176 strategies.

@@

Training of this strategy corresponds to finding maps from histories to
a probability of cooperation. The library includes a strategy trained
with $n_1 = m_1 = m_2 = 2$ that is \emph{mostly deterministic}, with 52 of the 64
probabilities being 0 or 1.

@@

Training of this strategy corresponds to finding parameters of the neural
network. An inner layer with just five nodes performs quite well in both deterministic and
noisy tournaments. The output of the ANN used in this work is deterministic;
a stochastic variant that outputs probabilities rather than exact moves could
be created.

@@

\begin{figure}[!hbtp]
    \centering
        \includegraphics[width=.5\textwidth]{FSM16.pdf}
        \caption{\bf Evolved\_FSM\_16: trained to maximize score in a standard
        tournament}
        \label{fig:fsm16}
\end{figure}

\begin{figure}
        \centering
        \includegraphics[width=.5\textwidth]{FSM16noise.pdf}
        \caption{\bf Evolved\_FSM\_16\_Noise\_05: trained to maximize score in a
        noisy tournament}
        \label{fig:fsm16noise}
\end{figure}

@@

There are several strategies based on ensemble methods that
are common in machine learning called Meta strategies. These strategies are
composed of a team of other strategies. In each round, each member of the team
is polled for its desired next
move. The ensemble then selects the next move based on a rule, such as the
consensus vote in the case of MetaMajority or the best individual performance
in the case of MetaWinner. These strategies were among the highest performing in 
the library before the inclusion of those trained by reinforcement learning.

@@

The trained strategies (denoted by a \(^{*}\) in
Appendix~\ref{app:list_of_players}) were trained using reinforcement
learning algorithms. The ideas of reinforcement learning can be attributed to
the original work of~\cite{turing1950computing} in which the notion that
computers would learn by taking random actions but according to a distribution
that picked actions with high rewards more often. The two particular algorithms
used here:

@@

Finally, Table~\ref{tbl:standard_ranks} and
Fig~\ref{fig:standard_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. Whilst there is some
stochasticity, the top three strategies almost always rank in the top three. For
example, the worst that the EvolvedLookerUp\_2\_2\_2 ranks in any tournament
is 8th.

@@

Results from noisy tournaments in which there is a 5\% chance that an action is
flipped are now described. As shown in Table~\ref{tbl:noisy_score} and
Fig~\ref{fig:noisy_score}, the best performing strategies in median payoff
are DBS, designed to account for noise, followed by two strategies trained in
the presence of noise and three trained strategies trained without noise. One of
the strategies trained with noise (PSO Gambler) actually performs less well than
some of the other high ranking strategies including
Spiteful TFT (TFT but defects indefinitely if the opponent defects twice
consecutively) and OmegaTFT (also designed to handle noise). While DBS is the clear
winner, it comes at a 6x increased run time over Evolved FSM 16 Noise 05.

@@

Fig~\ref{fig:comparison_cooperation_heatmaps_noisy_1}-~\ref{fig:comparison_cooperation_heatmaps_noisy_3} shows the rate of
cooperation in each round for the top three strategies (in the absence of noise)
and just as for the top performing strategies in the standard tournament
it is evident that
the strategies never defect first and learn to quickly punish poorer strategies.

@@

The tournament results indicate that pre-trained strategies are generally better
than human designed strategies at maximizing payoff against a diverse set of
opponents. An evolutionary algorithm produces strategies based on multiple
generic archetypes that are able to achieve a higher average
score than any other known opponent in a standard tournament. Most of the trained
strategies use multiple rounds of the history of play (some using all of it) and
outperform memory-one strategies from the literature. Interestingly, a trained
memory one strategy produced by a particle swarm algorithm performs well, better
than human designed strategies such as Win Stay Lose Shift and zero determinant
strategies (which enforce a payoff difference rather than maximize total payoff).

In opposition to historical tournament results and community folklore,
our results show that complex strategies can be effective for the
IPD\@. Of all the human-designed
strategies in the library, only DBS consistently performs well, and it is
substantially more complex than traditional tournament winners like TFT, OmegaTFT,
and zero determinant strategies.

The generic structure of the trained strategies did not appear to be
critical for the standard tournament -- strategies based on lookup tables,
finite state machines, neural networks, and stochastic variants all performed well.
Single layer neural networks performed well in both
noisy and standard tournaments
though these had some aspect of human involvement in the selection of features.
This is in line with the other strategies also where some human decisions are made
regarding the structure. For the LookerUp and Gambler archetypes
a decision has to be made
regarding the number of rounds of history and initial play that are to be used.
In contrast, the finite state machines and hidden Markov models
 required only a choice of the number
of states, and the training algorithm can eliminate unneeded states in the case
of finite state machines (evidenced by the unconnected nodes in the diagrams
for the included representations).

@@

Two strategies designed specifically to account for noise,
DBS and OmegaTFT, perform well and only DBS performs better than the trained
strategies and \textbf{only} in noisy contexts. Empirically we find that
DBS (with its default parameters) does not win tournaments at 1\% noise.
However DBS has a parameter that accounts for the expected amount of noise and a
followup study with various noise levels could make a more complete study of
the performance of DBS and strategies trained at various noise levels.
